An Analysis of the Use of Average Centipawn Loss to Detect Cheating in Chess

### 2022-12-09 Manuscript/manuscript.tex
added revised discussion

Evolutions in chess computing have allowed for extensive libraries of past games to be analyzed again utlizing centipawn math. A recent scandal involving the alleged cheating of Chess Grandmaster Hans Niemann in the sphere of professional chess has launched a conversation about how cheaters can be detected statistically to the forefront of Academic Data Science. This paper will explore how one might detect cheating using average centipawn loss and whether these methods are reliable enough to become a standard in the chess proctors handbook.

@@

@@

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{figures/ELOdistr}
    \caption{From these histogram plots of playing ELO it can be seen that the sample for ELOs below 2300 is very low thereby explaining the high fluctuations in the above line plots.}
    \label{fig:elo_histplot}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{figures/masterELOdistr}
    \caption{}
    \label{fig:elogt23_histplot}
\end{figure}

@@

-	Place plot of new histogram here
-	To better identify local regions of higher incidence, the number of bins and bin widths have been computed using the Freedman-Draconis rule.

@@

\begin{figure}
    \centering
    \includegraphics{figures/whiteACPL}
    \caption{}
    \label{fig:white_acpl}
\end{figure}

\begin{figure}
    \centering
    \includegraphics{figures/blackACPL}
    \caption{}
    \label{fig:black_acpl}
\end{figure}

\begin{figure}
    \centering
    \includegraphics{figures/elodiff-higher}
    \caption{}
    \label{fig:elodiff-higher}
\end{figure}

\begin{figure}
    \centering
    \includegraphics{figures/elodiff-lower}
    \caption{}
    \label{fig:elodiff-lower}
\end{figure}

@@

To better identify local regions of higher incidence, the number of bins and bin widths have been computed using the Freedman-Draconis rule.
Collapsing over the observations in each discrete bin to plot an estimate of central tendency along with a confidence interval
After plotting the estimated central tendency of ELO score against ACPL, one can observe that the result follows expectations. That is to say, there appears to be a negative correlation between player ranking and ACPL. As discussed in the introduction section at length, it is expected that as the rank of a player increases, they tend to lose less average centipawns, in effect playing a “stronger” game according to an engine. Calculating the statistics of linearity between ELO and ACPL yields a Pearson correlation coefficient of about -0.312 for Black and -0.286 for White. These r values need to be calculated separately for each color since their respective ACPL's are closely related within each game, hence they are not independent. Both values reflect a weak to moderate linear relationship of ELO and ACPL. It is possible that a reason to explain why these relationships are not more linear is that when a higher ELO white player get paired against a higher ELO Black player there is greater difficulty for each to find the “strongest” moves available, which may affect their resulting ACPL's. 
To observe this, one can simply plot the distribution of the ELO difference and ACPL of the higher and lower rated players.
-	Place plot and observations here
-	one can observe that in the case of the lower rated player the ELO difference matters much more than for the higher ELO player, who seems to have more or less the same accuracy whether playing against a player with similar ELO or a weaker player.
But the main focus here lies on whether Niemann's performance indicated suspicious behavior which could, possibly, be instances where he had cheated. When Niemann was playing White, the central tendency of his ACPL never appears to move outside the confidence interval of the binned observations in a way that would suggest his playing far exceeded his ranking. On the contrary, there seem to be times in which his ACPL was higher than expected such that he was perhaps playing worse than how he “should have”. Only when he played Black was there a single instance of binned observations where his ACPL was under the lower bound of that bin's confidence interval. Even so, as one can see in the distribution of ELO differences and ACPL of higher and lower rated players, this may not necessarily be due to foul play but instead by simply being paired against a stronger/weaker opponent.

@@

% rewrite findings and improvments

>>>>

Evolutions in chess computing have allowed for extensive libraries of past games to be analyzed again utlizing centipawn math. A recent scandal involving the alleged cheating of Chess Grandmaster Hans Niemann in the sphere of professional chess has launched a conversation about how cheaters can be detected statistically to the forefront of Academic Data Science. This paper will explore how one might detect cheating using average centipawn loss and whether these methods are reliable enough to become standard in the chess proctors handbook.

@@

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{figures/ELOdistr}
    \caption{From these histogram plots of playing ELO it can be seen that the sample for ELOs below 2300 is very low thereby explaining the high fluctuations in the above line plot.}
    \label{fig:elo_histplot}
\end{figure}

@@

@@

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{figures/masterELOdistr}
    \caption{The filtered data uses a more normal dirstribution of ELO scores.}
    \label{fig:elogt23_histplot}
\end{figure}

To better identify local regions of higher incidence, the number of bins and bin widths have been computed using the Freedman-Diaconis rule.

\begin{equation}
    \label{eq:freedman-diaconis}
    BIN WIDTH = 2 \frac{iqr}{n^{1/3}},
\end{equation}

where iqr is the interquartile range of the data and n is observations.

@@

@@

After plotting the estimated central tendency of ELO score against ACPL, one can observe that the result follows expectations. That is to say, there appears to be a negative correlation between player ranking and ACPL.

\begin{figure}[!htb]
    \centering
    \includegraphics{figures/whiteACPL}
    \caption{Relationship of Niemann's ELO score and ACPL while playing as White.}
    \label{fig:white_acpl}
\end{figure}

\begin{figure}[!htb]
    \centering
    \includegraphics{figures/blackACPL}
    \caption{Relationship of Niemann's ELO score and ACPL while playing as Black.}
    \label{fig:black_acpl}
\end{figure}

As discussed in the introduction section at length, it is expected that as the rank of a player increases, they tend to lose less average centipawns, in effect playing a “stronger” game according to an engine. Calculating the statistics of linearity between ELO and ACPL yields a Pearson correlation coefficient of about -0.312 for Black and -0.286 for White. These r values need to be calculated separately for each color since their respective ACPL's are closely related within each game, hence they are not independent. Both values reflect a weak to moderate linear relationship of ELO and ACPL. It is possible that a reason to explain why these relationships are not more linear is that when a higher ELO white player get paired against a higher ELO Black player there is greater difficulty for each to find the “strongest” moves available, which may affect their resulting ACPL's. To observe this, one can simply plot the distribution of the ELO difference and ACPL of the higher and lower rated players.

\begin{figure}[!htb]
    \centering
    \includegraphics{figures/elodiff-higher}
    \caption{Distribution of ELO difference and ACPL of the higher rated players in the dataset.}
    \label{fig:elodiff-higher}
\end{figure}

\begin{figure}[!htb]
    \centering
    \includegraphics{figures/elodiff-lower}
    \caption{Distribution of ELO difference and ACPL of the lower rated players in the dataset.}
    \label{fig:elodiff-lower}
\end{figure}

One can observe that in the case of the lower rated player the ELO difference matters much more than for the higher ELO player, who seems to have more or less the same accuracy whether playing against a player with similar ELO or a weaker player.

But the main focus here lies on whether Niemann's performance indicated suspicious behavior which could, possibly, be instances where he had cheated. When Niemann was playing White, the central tendency of his ACPL never appears to move outside the confidence interval of the binned observations in a way that would suggest his playing far exceeded his ranking. On the contrary, there seem to be times in which his ACPL was higher than expected such that he was perhaps playing worse than how he “should have”. Only when he played Black was there a single instance of binned observations where his ACPL was under the lower bound of that bin's confidence interval. Even so, as one can see in the distribution of ELO differences and ACPL of higher and lower rated players, this may not necessarily be due to foul play but instead by simply being paired against a stronger/weaker opponent.

@@

The purpose of this research was meant to analyze how one could detect foul play in chess by observing the correlation between a player's rating and ACPL over a period of time; but rather than painting a clearer picture of what many consider to be a common method to identify cheaters, more questions have been raised than answers. In the case of Hans Niemann, evidently the data does not even suggest a strong linear relationship between his ELO score and ACPL over time. Upon further research, it appears that the claims made by many regarding the existence of this relationship is somewhat contested. Regrettably, the only real contribution this paper provides is that, in the case of Hans Niemann's professional history, it is inconclusive whether the methodologies used here can even be used to accurately indicate “cheating”. The fact that the data used in this paper only reflects the history of a single professional player may be a significant limitation in investigating the true relationship of these variables. Shifting the focus from Niemann and performing the same analysis on a dataset consisting of players that were randomly selected from another database may provide more accurate results of what one could expect this relationship could be for the average player. Another definite limitation of this study came about as a result during the process of binning observations before plotting the central tendency of ELO score against ACPL. It is possible that during this process, the visibility of certain observations that could have been suspicious would undoubtedly be lessened when binned with others that were less so. This would have effectively drowned out any times where Niemann could possibly have “cheated” inconsistently. Which leads to undoubtedly the biggest limitation of this study, possible inability to detect certain styles of cheating. Although it might be possible to detect a consistent trend where an individual may be playing far better than expected for their ranking by just observing the relationship between ELO score and ACPL, this may not be so for an inconsistent or “smart” cheater. A “smart” cheater might prevent themselves from relying on moves suggested by an engine too regularly and may, instead, elect to make only certain moves that were suggested. This would only give them a slight advantage, but it would be nearly impossible to detect using a simple correlation. The possible future directions of this study at the moment would be to investigate this same relationship between ELO score and ACPL for other professional chess players to observe if the same or different effects could be observed.
